---
title: "Week 11 Bayes Continued"
author: "Lida Bentz"
date: "11/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Bayes: Is our coin fair?

we flip the coin and get $H$. 

Q: What can we say about whether the coin is fair or not fair. 

Having the Bayes tool we decide to apply that. 

We need a set of hypothesis. $p=P(H)$. We don't know $p$ _but we can model things we don't know with random variables_

What possible values could $p$ take? $0\leq p \leq 1$ 

We can come up with a __discrete__ model where we choose finitely many values for $p$ and assign a value to $P(p=x)$ for each value, with the constraint that $\sum_x P(p=x)=1$. 

We could do a __continuous__ model any $p=x$ between $0\leq x\leq 1$ is possible and assign a density $f(x)$ where density is $\int_a^b f(x) dx=P(a\leq p\leq b)$. 

## Discrete Model
We need to choose discrete values as hypothesis.
```{r}
x=seq(from=0, to=1,by=0.01)
x
```
Above are our hypothesis.
Below is our prior.
```{r}
hp=1/length(x)
hp
```
According to Bayes, given a value of $p$ how likely is it that I have flipped the coin and gotten $H$, so $P(H_p)=x$
```{r}
PH_p=x
```
Then add up my prior estimates
```{r}
tot=sum(hp*PH_p)
Pp_H=hp*PH_p/tot
Pp_H
sum(Pp_H)
```

So we plot them
```{r}
plot(x, Pp_H)
```
So one coin flip is not that much data. Let's flip it again, we get a H (heads). Given a new coin flip, we use $PH_p$ as the new prior estimate. 
```{r}
prior=Pp_H
P2Hp=x
tot=sum(x*prior)
posterior=x*prior/tot

plot(x,posterior)
```
### flip the coin a third time. We get a T
```{r}
prior=posterior
HT_p=(1-x)
tot=sum(HT_p*prior)
posterior=HT_p*prior/tot

plot(x, posterior)
```
 The finer you make the discrete model, it will converge to the continuous model. 
## Continuous Model
 In the continuous case, the prior is some valid density funciton $f(x)$. A valid density is if we integrate over the density, we get valid probability. This implies
 $$f(x)>0$$
 $$\int_0^1 f(x) dx=1$$
 Any $f$ that satisfies the above is a possibility. There is a whole family of functions call beta distributions that meet this criteria. 
 A flat prior would be a constant density, $f(x)=1$. or $f(x)=2x$. 
 Then we need to assign $P(H\vert p=x)=x$. (assign for $f(x)=1$). Then 
 $$totj=\int_0^1 P(H\vert p=x)\cdot f(x) dx=\int_0^1 x dx=\frac{1}{2}$$
 Then $$P(p=x\vert H)=\frac{x}{\frac{1}{2}}=2x$$
 
 
### Suppose we flip it again 
 and get a heads, 
 $P(H\vert p=x, H)$ still is $x$. 
 Now $tot=\int_0^1 x\cdot 2x dx=\frac{2}{3} x^3$. Thus my new posterior is $P(HH\vert p=x)=\frac{2x^2}{\frac{2}{3}}=3x^2$. 
 
### Third flip 
 is a tail
 $P(T\vert p=x, HH)=1-x$. 
 $$tot=\int_0^1 (1-x)3x^2 dx=\int_0^1 (3x^2-3x^3) dx=1-\frac{3}{4}=1/4$$
 THEN $$P(HHT\vert p=x)=\frac{(1-x)3x^2}{1/4}=12(1-x)x^2$$

Some notes that THC at 20% more lucrative than lower. How would you prove the suspition that the lab is manipulating to get values over 20%. _Given a set of concentration measurements, can we detect if the lab manipulated them?_
What you will need to do to answer this question, you will need a model aka some hypothesis for distributions. 
 